D3064R0: OOTA Execution on Real Implementations is Provably Vacuous

### 2023-12-16 ootavacuous.tex
Document user influence over semantics
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@

The exact definition of a computer language is subject to some debate,
with standards, implementations, and users all having some degree of
influence~\cite{KayvanMemarian2016DepthOfC-1,KayvanMemarian2016DepthOfC-2},
and each of which is subject to change over time.
Nevertheless, because volatile atomics constitute observable
behavior~\cite[\co{intro.abstract}]{ThomasKoeppe2023N4950},
they must be executed in strict accordance with the rules of the abstract
machine.
This means that any volatile atomic operation involving a store
must execute as if all loads whose return values are used to compute
that store's address or value have already returned.
OOTA afficienados will recognize this as a special case of ordering
relaxed loads before relaxed stores, albeit one not requiring
expensive memory-fence instructions on weakly ordered architectures.

Although C++ non-volatile accesses to atomic objects are not observable
behavior, they are still subject to the memory model.
In particular, implementations must not behave as if the source code has
additional:

@@

@@

>>>>

\subsubsection{Implementers and Users Influence ``Semantic Dependencies'' Definition}
\label{sec:Implementers and Users Influence ``Semantic Dependencies'' Definition}

The exact definition of a computer language is subject to some debate,
with standards, implementations, and users all having some degree of
influence~\cite{KayvanMemarian2016DepthOfC-1,KayvanMemarian2016DepthOfC-2},
and each of which is subject to change over time.
Users  and implementers are of course wise to assume that their opinions
might be overridden by those of the standard and, for the users,
especially the implementation.
However, consider the following code:

\begin{quote}
	\co{void foo(char c} \\
	\co{\{} \\
	\co{\ \ \ \ return c + 128;} \\
	\co{\}}
\end{quote}

The standard does not specify whether or not type \co{char} is signed,
leaving that choice to the implementations, which follow the lead
of the architects of a given CPU family.
Therefore, the code above is undefined behavior on systems choosing
signed \co{char} and having 8-bit bytes (so that there is no semantic
dependency from the \co{foo()} function's argument to its return value),
but perfectly well-defined on systems having unsigned \co{char}
(so that there is a semantic dependency).
So score one for the implementers, albeit by conscious choice on the
part of the standards committee.

Except that GCC provides the \co{-funsigned-char} command-line
argument that causes this implementation to treat variables of
type \co{char} as unsigned, regardless of the architects' wishes.
Given this command-line argument, there is always a semantic dependency
from the \co{foo()} function's argument to its return value.
So score one for at least some implementations' users.

Please see Appendix~\ref{sec:User Influence Over Language Semantics}
for more examples of user control over semantics, including semantic
dependencies.

@@

Because volatile atomics constitute observable
behavior~\cite[\co{intro.abstract}]{ThomasKoeppe2023N4950},
they must be executed in strict accordance with the rules of the abstract
machine.
This means that any volatile atomic operation involving a store
must execute as if all loads whose return values are used to compute
that store's address or value have already returned.
OOTA afficienados will recognize this as a special case of ordering
relaxed loads before relaxed stores, albeit one not requiring
expensive memory-fence instructions on weakly ordered architectures.

Although C++ non-volatile accesses to atomic objects are not observable
behavior, @@@

In addition,such accesses are still subject to the memory model.
In particular, implementations must not behave as if the source code has
additional:

@@

\clearpage

@@

\clearpage

\section{User Influence Over Language Semantics}
\label{sec:User Influence Over Language Semantics}

As noted earlier, the exact definition of a computer language is subject
to some debate, with standards, implementations, and users all having
some degree of
influence~\cite{KayvanMemarian2016DepthOfC-1,KayvanMemarian2016DepthOfC-2},
and each of which is subject to change over time.
It is natural to dismiss user influence when compared to the text of
standards or the code in implementations, but both are subject to
change and do change over time.
The implementation is especially subject to change, for example,
consider the following command line used to build the Linux kernel's
\co{kernel/rcu/tree.c} C-language source file:

~\\
{
	\scriptsize
	\texttt{
	gcc -Wp,-MMD,kernel/rcu/.tree.o.d -nostdinc
	-I./arch/x86/include -I./arch/x86/include/generated  -I./include
	-I./arch/x86/include/uapi -I./arch/x86/include/generated/uapi
	-I./include/uapi -I./include/generated/uapi
	-include ./include/linux/compiler-version.h
	-include ./include/linux/kconfig.h -include
	./include/linux/compiler\_types.h -D\_\_KERNEL\_\_
	-fmacro-prefix-map=./= -Werror -std=gnu11 -fshort-wchar
	-funsigned-char -fno-common -fno-PIE -fno-strict-aliasing -mno-sse
	-mno-mmx -mno-sse2 -mno-3dnow -mno-avx -fcf-protection=branch
	-fno-jump-tables -m64 -falign-jumps=1 -falign-loops=1
	-mno-80387 -mno-fp-ret-in-387 -mpreferred-stack-boundary=3
	-mskip-rax-setup -mtune=generic -mno-red-zone -mcmodel=kernel
	-Wno-sign-compare -fno-asynchronous-unwind-tables
	-mindirect-branch=thunk-extern -mindirect-branch-register
	-mindirect-branch-cs-prefix -mfunction-return=thunk-extern
	-fno-jump-tables -fpatchable-function-entry=16,16
	-fno-delete-null-pointer-checks -O2 -fno-allow-store-data-races
	-fstack-protector-strong -fomit-frame-pointer
	-fno-stack-clash-protection -falign-functions=16
	-fno-strict-overflow -fno-stack-check -fconserve-stack
	-Wall -Wundef -Werror=implicit-function-declaration
	-Werror=implicit-int -Werror=return-type -Werror=strict-prototypes
	-Wno-format-security -Wno-trigraphs -Wno-frame-address
	-Wno-address-of-packed-member -Wframe-larger-than=2048 -Wno-main
	-Wno-unused-but-set-variable -Wno-unused-const-variable -Wvla
	-Wno-pointer-sign -Wcast-function-type -Wno-array-bounds
	-Wno-alloc-size-larger-than -Wimplicit-fallthrough=5
	-Werror=date-time -Werror=incompatible-pointer-types
	-Werror=designated-init -Wenum-conversion
	-Wno-unused-but-set-variable -Wno-unused-const-variable
	-Wno-restrict -Wno-packed-not-aligned -Wno-format-overflow
	-Wno-format-truncation -Wno-stringop-overflow
	-Wno-stringop-truncation -Wno-missing-field-initializers
	-Wno-type-limits -Wno-shift-negative-value
	-Wno-maybe-uninitialized -Wno-sign-compare
	-DKBUILD\_MODFILE='"kernel/rcu/tree"' -DKBUILD\_BASENAME='"tree"'
	-DKBUILD\_MODNAME='"tree"' -D\_\_KBUILD\_MODNAME=kmod\_tree -c
	-o kernel/rcu/tree.o kernel/rcu/tree.c
	}
}

We do not propose to explain all of these, and sufficiently motivated
readers can avail themselves of the GCC documentation.
We instead look at representative members of several categories.

The \co{-funsigned-char} causes type \co{char} to be unsigned, which
overrides per-architecture defaults, some of which treat \co{char}
as signed and others as \co{unsigned}.
This choice prevents a class of bugs, and also allows the kernel to make
reliable use of the uppermost bit of variables of type \co{char}.
It also affects the definition of ``semantic dependency'' by changing
the arithmetic proporties of this type.
In theory, the standard could specify the signedness of \co{char},
but presumably the variety of existing practice in the 1980s prevented
this from being done.

The \co{-mno-sse} prevents GCC from making use of the SSE hardware.
This is done for performance reasons, as it avoids the overhead of
saving and restoring the state of this hardware when switching between
user and kernel contexts.
Similarly, the \co{-mcmodel=kernel} causes the kernel binary to be
placed in the uppermost 2GB of the address space, again reducing
the overhead of switching between user and kernel contexts.
These are cases where the standard does not specify anything, nor
should it.

The \co{-fpatchable-function-entry=16,16} causes GCC to emit 16
\co{nop} instructions at the beginning of each function, with
the function's entry point being just after this string of \co{nop}
instructions.
The resulting buffers are used by the Linux-kernel tracing infrastructure
which is in turn used for debugging, performance measurement, and
monitoring.
This is again clearly outside of the scope of the standard.

The \co{-fstack-protector-strong} causes GCC to emit code that provides
some protection against some clases of attacks based on buffer overflows.
One could rightly argue code should simply avoid ever overflowing buffers,
but things like memory allocators and userspace memory accesses must
use code that can be difficult to distinguish from buffer overflows.
It is not clear that the ever-increasing variety of attacks should
affect the standard.

The \co{-fno-strict-overflow} causes GCC to act as if signed integer
overflow is defined behavior.
Note well that this command-line option also affects the definition of
``semantic dependency''.
This might be a controversial choice, and another option would be to add
a new set of signed integer types to the standard for which overflow is
defined as wrapping, similar to the situation with unsigned integers.

The \co{-Werror=strict-prototypes} causes GCC to warn if old-style
non-ANSI function prototypes are used.
This helps avoid certain classes of bugs.
Warnings are by design outside of the scope of the standard.

In other words, user preference can exert a non-trivial influence over
language semantics, and in particular can affect some aspects of the
definition of ``semantic dependency''.
